DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2021-10-30 manuscript/manuscript.Rmd
results

```{r analysis-preferences}
# Seed for random number generation
set.seed(50)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
data(opentree_chronograms)
rpackage_available <- FALSE
availability <- ifelse(rpackage_available, "is available", "will be available")
```

@@

# Description

The `datelife` workflow builds off from functions in several R packages (rotl [@Michonneau2016], ape [@Paradis2004],
geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], and phylocomr [@Ooms2018]).

@@

# Example
We used `r cite_r("r-references.bib")` for all our analyses.

>>>>

```{r analysis-preferences}
# Seed for random number generation
set.seed(50)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
data(opentree_chronograms)
rpackage_available <- FALSE
availability <- ifelse(rpackage_available, "is available", "will be available")
dq <- datelife::make_datelife_query(input = "fringillidae", get_spp_from_taxon = TRUE)
```

@@

# Implementation/Description/Workflow

The `datelife` workflow builds off of functions from several R packages (rotl [@Michonneau2016], ape [@Paradis2004],
geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], and phylocomr [@Ooms2018]).

@@

# Results

<!--We used `r cite_r("r-references.bib")` for all our analyses.-->

## Case study

We illustrate the `datelife` workflow using the family of true finches, Fringillidae as an example. To contextualize, a college educator wishes to know the state-of-the-art on time of evolutionary origin of species belonging to the true finches. 
One option is to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills. 
The first step is to run a higher-taxon-name 'datelife` query. This will get taxon names for all recognised species within any higher taxon. The Fringillidae has `r length(dq$cleaned_names)` species, according to the Open Tree of Life taxonomy.
Once with a curated set of query taxon names, the next step is to run a `datelife` search. This will find all chronograms that contain at least two queried taxon names, and will save the information on time of lineage divergence as (an R "data frame") table.
 There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig. \ref{fig:schronograms1}).
<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig. \ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche]. -->
The final step is to summarize the available information using the two alternative types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages are used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:summaries}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM. With both methods, ages are generally consistent with source ages, but there are some biological examples in which this is not true (see Discussion).
<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

## Cross-validation test
Data from source chronograms can be also used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs. \ref{fig:cvbladj} and \ref{fig:cvbold}). As a form of cross validation, we took tree topologies from each study and calibrated them using time of lineage divergence data from all other source chronograms. In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
<!-- We need to think about the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
We also demonstrate the usage of PATHd8 [@Britton2007] as an alternative method to BLADJ. For this, we run a `datelife` branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System [BOLD; ratnasingham2007bold] to generate branch lengths. We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms. However, dating with PATHd8 using congruified calibrations, was only successful in
<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}). From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD  data for some species is absent from the database. Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.

### 2021-10-30 manuscript/paper_3-main.Rmd
results

test improvement in synthetic otol tree after using bold sequences. Is it really better resolved? Probably not.
Do not include this analysis in paper, maybe in examples, but i think it is not worthy.
-->
In this section we demonstrate the types of outputs that can be obtained with `datelife`, using as an example the bird family Fringillidae of true finches. We performed a higher-taxon search to obtain data on lineage divergence available in `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` species according to the Open Tree of Life taxonomy). There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig. \ref{fig:schronograms1}).
<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig. \ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche]. -->
Data from these source chronograms was used to generate two types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms was first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages were used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:summaries}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM. With both methods, ages are generally consistent with source ages, but there are some biological examples in which this is not true (see Flaws, Limitations and Prospects section).
<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs. \ref{fig:cvbladj} and \ref{fig:cvbold}). As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms. In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
Maximum tree ages were only approximately recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
<!-- We need to think abput the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
To exemplify dating a tree with branch lengths provided by the user, we obtained DNA sequence data using the Barcode of Life Data System [BOLD; ratnasingham2007bold] and branch lengths were successfully generated for all source chronograms. However, dating with PATHd8 [@Britton2007] using congruified calibrations, was only successful in
<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}). From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD. Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developes on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.

>>>>

test improvement in synthetic otol tree after using bold sequences. Is it really better resolved? Probably not.
Do not include this analysis in paper, maybe in examples, but i think it is not worthy.
-->
